Towards detecting critical crowd density with Wi-Fi positioning: Case Amsterdam ArenA

### 2017-01-23 PaperOnSmoothing/draft-31052016.tex
changes in paper

The RSS value of a signal decreases when the distance between transmitter and receiver increases.
The decrease is given by a path loss model, given by the Friis equation
\begin{equation}
P_{R}=P_{T}+10\lambda\log_{10}(\frac{c}{4\pi fr})
\label{friis}
\end{equation}
where we use $\lambda=2$ (line of sight condition), $c$ is the speed of light, $f$ is 2.4 GHz, and $r$ is the distance between the mobile device and the AP.\\
The measured RSS values contain unpredictable variation due to noise, and interference such as absorption and reflection by obstacles between transmitter and receiver (shadowing and multipath effects) \cite{goldsmith}. The positions of these obstacles may change in time, in particular in the case of human bodies.
As a result, the circles in the lateration method do not intersect at a unique point (see e.g. Fig.\ref{fig:trilateration}). 
In this case an optimization procedure is undertaken for positioning of the smart phone.
We use the least squares optimization method, which in our case takes the form of a chi-square data fit. Exact description of the method is beyond the scope of this paper and we refer the reader to \cite{bevington} for details. We note, however, that the statistical estimation of the position provides as also with standard deviation $\sigma$ of the estimation, which we will use in Sec. \ref{sec:density}.
\begin{comment}
Based on the chi-square fit, uncertainty values for the parameter estimates are calculated.
These values represent parameter uncertainties that correspond to a $1\sigma$ error or an increase of $\chi^2+1$. 
The $1\sigma$ error values form the input to the smoothing method of the density estimation discussed in this paper.
\end{comment}

@@

\item {\it Issue 2: volatility of packet rates}. When a MAC device is connected to the Wi-Fi internet, it sends packets with a relatively stable and frequent rate. However, during concert hours, very few devices are actually connected to the internet. So, most of our estimations of positions come from signals that the device sends in a "probing" mode, i.e. while searching for a network. In this case the packet rate is quite volatile, ranging from a few seconds to a few minutes (citation?)\marginpar{add citation}. This means that when we make a snapshot of the MAC addresses visible in a certain moment, we are detecting a only a very small fraction of the MAC devices.

\item {\it Issue 3: MAC address randomization}. Due to ever increasing privacy concerns and possibly other business reasons, starting from 2014, the Apple I-phones have introduced randomization of the MAC address when the device is in a probing mode (not connected to internet) (citation?)\marginpar{add citation}. This means that when the device switches from a probing to an online mode or vice versa, the MAC address suddenly changes and thus the device cannot be followed anymore. It is worth noting, however, that from the MAC address itself it can be determined whether the address is an original MAC address (non-randomized) or randomized. 
\end{enumerate}

@@

In order to deal with issue 1 from above, let us start with the following observation: we are not interested in the individual locations of the MAC devices, but rather in the density of the crowd. Also, we would  especially like to have more precise estimation when the density reaches a certain threshold, so that the crowd outside the dense region can be promptly navigated away from it. 
This situation (dense crowd), means that there is more blockage and absorption of signals from bodies. This leads to less precise estimation of the distance between a MAC device and a Wi-Fi router based on Wi-Fi signal strength, which means that the rings in e.g.  Fig.\ref{fig:trilateration_problem} would be even thicker. This exaggerates the effect of "bimodal" or even multi-modal distribution of position estimates. Thus, in a dense crowd, the best that we can derive from the positioning method for a MAC device is a probability distribution over all possible locations. For example, in the scenario in Fig. \ref{fig:trilateration_problem}, we can say that the MAC device is with a probability of 0.5 in the left grey region (region where all three rings overlap) and with a probability of 0.5 in the right grey region \footnote{In this example we assign the probabilities in a trivial way for the purpose of demonstrating our idea}. While this does not provide us with  very useful information about the location of the MAC device, if we apply the same reasoning for all MAC devices, and we add together the spatial probability density functions for all MAC devices, we end up with a spatial distribution of the crowd density. If we assume that the locations of the MAC devices are mutually independent and identically distributed, we can apply directly the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per square meter will vanish. However, we cannot assume the mentioned criteria,  because people tend to go to concerts in groups. In this case, the variance of the estimation in the limiting case is equal to the average correlation between the locations. Nevertheless, we can safely assume that for a regular crowd at a concert the group size is relatively small compared to the whole crowd and thus the average correlation tends to zero as the number of people increases. Thus, the error of the estimation of the crowd density will in anyway diminish as the number of people grows.

In what follows we formalize the discussion above. 
Let $\{mac_1, mac_2,...mac_n\}$ be all MAC devices detected at time $t$. Let $R$ be an arbitrary region from the stadium.  Let $X_i$ be a random variable defined by

@@

It is reasonable to assume that there is an upper limit $\lambda$ of a number of people that go to a concert together (a group), that is, whose locations are correlated\footnote{Actually, when the crowd is so dense that people can not move freely anymore, the whole crowd becomes a "group" (as in the  Love Parade disaster) and all locations are correlated. But we aim to detect a high density with our method {\it before} this happens, to react preventively; otherwise, it is too late.}. Note that because the random variables $\{X_i\}_{i=1}^n$ take values in $\{0,1\}$, the covariances $Cov(X_i,X_j)$ take values in $[-1,1]$. Thus,  the covariances are right-bounded (by $1$). Denote by $\kappa$ the maximal covariance between any $X_i$ and $X_j$ and let us write $i \sim j$ if and only if the owners of the MAC devices $mac_i$ and $mac_j$ are in the same group of friends. 
Then, 
\begin{equation}\label{eq_covar}
\sum_{i \not =  j}Cov(X_i,X_j) = 2 \sum_{1\leq i <  j \leq n }Cov(X_i,X_j) =  2 \sum_{i,j: i \sim j }Cov(X_i,X_j)  \leq 2 n \cdot \frac{\lambda(\lambda - 1)}{2} \cdot\kappa = n  \kappa \lambda(\lambda - 1)
\end{equation} 
where the inequality holds because the maximal number of groups is $n$ and the maximal number of pairs $(i,j)$ in a group is $\lambda(\lambda-1)/2$.
Let $$\mu = \frac{1}{n}\sum_{i=1}^{n} Var(X_i),$$ that is, 
denote by $\mu$ the average variance of ${X_1, X_2, ..., X_n}$ (note that $\mu < 1$ from the definition of $\{X_i\}_{i=1}^n$ ).  From \eqref{eq_Var} and \eqref{eq_covar} we have

\begin{equation}
Var(\frac{X}{n} ) \leq \frac{1}{n^2}(n \mu + n  \kappa \lambda(\lambda - 1)) = \frac{1}{n}(\mu +  \kappa \lambda(\lambda - 1)), 
\end{equation}
which tends to $0$ when $n \rightarrow \infty$. Note that we have greatly overestimated the covariance with the inequation in \eqref{eq_covar}, which means that in practice the variance converges to $0$ much faster than as presented.

@@

\subsubsection{MAC address randomization: using Big data again}

To address the last issue, {\it MAC address randomization}, we rely once again on the fact that we have a lot of data. Namely, we know from the structure of the MAC address whether it has been randomized or not. Figure \ref{fig:randomized} shows the ratio between randomized and non-randomized addresses observed per minute from midnight until 6:00 am. We can see that it is quite stable through time; in fact, the Pearson correlation coefficient between the time series of randomized and non-randomized addresses is 0.88.

\begin{figure}[h!]
	\centering
	\includegraphics[width=130mm]{nonrandomized.jpeg}
	\caption{Non-randomized addresses observed per minute}
	\label{fig:nonrandomizedadresses}
\end{figure} 
\begin{figure}[h!]
	\centering
	\includegraphics[width=130mm]{randomized.jpeg}
	\caption{Randomized addresses observed per minute}
	\label{fig:randomizedaddresses}
\end{figure} 
\begin{figure}[h!]
	\centering
	\includegraphics[width=130mm]{RatioRandNonrand.jpeg}
	\caption{Ratio between randomized and non-randomized addresses observed per minute}
	\label{fig:randomized}
\end{figure}

\noindent Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by 1.225, to account for the ignored MAC devices. We choose factor 1.225 to allow for a safety margin: in the peaks of the graph in Fig. \ref{fig:randomized} the values are 0.225, and we would rather overestimate than underestimate the crowd density. \marginpar{in general: calculate it automatically, with a confidence interval}
Note that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market. 
\marginpar{mention the absolute numbers, or give graphs}

@@

>>>>

The RSS value of a signal decreases when the distance between transmitter and receiver increases.
The decrease is given by a path loss model, given by the Friis equation
\begin{equation}
P_{R}=P_{T}+10\lambda\log_{10}(\frac{c}{4\pi fr})
\label{friis}
\end{equation}
where we use $\lambda=2$ (line of sight condition), $c$ is the speed of light, $f$ is 2.4 GHz, and $r$ is the distance between the mobile device and the AP.\\
The measured RSS values contain unpredictable variation due to noise, and interference such as absorption and reflection by obstacles between transmitter and receiver (shadowing and multipath effects) \cite{goldsmith}. The positions of these obstacles may change in time, in particular in the case of human bodies.
As a result, the circles in the lateration method do not intersect at a unique point (see e.g. Fig.\ref{fig:trilateration}). 
In this case an optimization procedure is undertaken for positioning of the smart phone.
We use the least squares optimization method, which in our case takes the form of a chi-square data fit. Exact description of the method is beyond the scope of this paper and we refer the reader to \cite{bevington} for details. We note, however, that the statistical estimation of the position provides us also with standard deviation $\sigma$ of the estimation, which we will use in Sec. \ref{sec:density}.
\begin{comment}
Based on the chi-square fit, uncertainty values for the parameter estimates are calculated.
These values represent parameter uncertainties that correspond to a $1\sigma$ error or an increase of $\chi^2+1$. 
The $1\sigma$ error values form the input to the smoothing method of the density estimation discussed in this paper.
\end{comment}

@@

\item {\it Issue 2: volatility of packet rates}. When a MAC device is connected to the Wi-Fi internet, it sends packets with a relatively stable and frequent rate. However, during concert hours, very few devices are actually connected to the internet. So, most of our estimations of positions come from signals that the device sends in a "probing" mode, i.e. while searching for a network. In this case the packet rate is quite volatile, ranging from a few seconds to a few minutes (citation?)\marginpar{add citation}. This means that when we make a snapshot of the MAC addresses visible in a certain moment, we are detecting only a very small fraction of the MAC devices.

\item {\it Issue 3: MAC address randomization}. Due to ever increasing privacy concerns and possibly other business reasons, starting from 2014, the Apple I-phones have introduced randomization of the MAC address when the device is in a probing mode (not connected to internet) (citation?)\marginpar{add citation}. This means that devices not connected to the internet are continuously changing their MAC address and canot be followed over time.  It is worth noting, however, that from the MAC address itself it can be determined whether the address is an original MAC address (non-randomized) or randomized. 
\end{enumerate}

@@

In order to deal with the bimodal distribution issue  from above,  let us start with the following observation: we are not interested in the individual locations of the MAC devices, but rather in the density of the crowd. Also, we would  especially like to have more precise estimation when the density reaches a certain threshold, so that the crowd outside the dense region can be promptly navigated away from it. 
This situation (dense crowd), means that there is more blockage and absorption of signals from bodies. This leads to less precise estimation of the distance between a MAC device and a Wi-Fi router based on Wi-Fi signal strength, which means that the rings in e.g.  Fig.\ref{fig:trilateration_problem} would be even thicker. This exaggerates the effect of "bimodal" or even multi-modal distribution of position estimates. Thus, in a dense crowd, the best that we can derive from the positioning method for a MAC device is a probability distribution over all possible locations. For example, in the scenario in Fig. \ref{fig:trilateration_problem}, we can say that the MAC device is with a probability of 0.5 in the left grey region (region where all three rings overlap) and with a probability of 0.5 in the right grey region \footnote{In this example we assign the probabilities in a trivial way for the purpose of demonstrating our idea}. While this does not provide us with  very useful information about the location of the MAC device, if we apply the same reasoning for all MAC devices, and we add together the spatial probability density functions for all MAC devices, we end up with a spatial distribution of the crowd density. If we assume that the locations of the MAC devices are mutually independent and identically distributed, we can apply directly the law of the large numbers and conclude that for a dense crowd the error of the estimation of the density per square meter will vanish. However, we cannot assume the mentioned criteria because people tend to go to concerts in groups, that is, their locations are correlated. Thus, the variance of the estimation in the limiting case is equal to the average correlation between the locations.
We will demonstrate that the average correlation still tends to zero as the number of people increases,  because for a regular crowd at a concert, the group size is relatively small compared to the whole crowd. Thus, the error of the estimation of the crowd density will in anyway diminish as the number of people grows.

We proceed with formalizing the discussion above. 
Let $\{mac_1, mac_2,...mac_n\}$ be all MAC devices detected at time $t$. Let $R$ be an arbitrary region from the stadium.  Let $X_i$ be a random variable defined by

@@

It is reasonable to assume that there is an upper limit $\lambda$ of a number of people that go to a concert together (a group), that is, whose locations are correlated\footnote{Actually, when the crowd is so dense that people can not move freely anymore, the whole crowd becomes a "group" (as in the  Love Parade disaster) and all locations are correlated. But we aim to detect a high density with our method {\it before} this happens, to react preventively; otherwise, it is too late. However, it is important as future work to study at which density groups start to merge, in light of studies such as \cite{density}.}. Note that because the random variables $\{X_i\}_{i=1}^n$ take values in $\{0,1\}$, the covariances $Cov(X_i,X_j)$ take values in $[-1,1]$. Thus,  the covariances are right-bounded (by $1$). Denote by $\kappa$ the maximal covariance between any $X_i$ and $X_j$ and let us write $i \sim j$ if and only if the owners of the MAC devices $mac_i$ and $mac_j$ are in the same group of friends. 
Then, 
\begin{equation}\label{eq_covar}
\sum_{i \not =  j}Cov(X_i,X_j) = 2 \sum_{1\leq i <  j \leq n }Cov(X_i,X_j) =  2 \sum_{i,j: i \sim j }Cov(X_i,X_j)  \leq 2 n \cdot \frac{\lambda(\lambda - 1)}{2} \cdot\kappa = n  \kappa \lambda(\lambda - 1)
\end{equation} 
where the inequality holds because the maximal number of groups is $n$ and the maximal number of pairs $(i,j)$ in a group is $\lambda(\lambda-1)/2$.
Let $$\mu = \frac{1}{n}\sum_{i=1}^{n} Var(X_i),$$ that is, 
denote by $\mu$ the average variance of ${X_1, X_2, ..., X_n}$ (note that $\mu \leq 1$ from the definition of $\{X_i\}_{i=1}^n$ ).  From \eqref{eq_Var} and \eqref{eq_covar} we have

\begin{equation}
Var(\frac{X}{n} ) \leq \frac{1}{n^2}(n \mu + n  \kappa \lambda(\lambda - 1)) = \frac{1}{n}(\mu +  \kappa \lambda(\lambda - 1)), 
\end{equation}
which tends to $0$ when $n \rightarrow \infty$. Note that we have greatly overestimated the covariance with the inequality in \eqref{eq_covar}, which means that in practice the variance converges to $0$ much faster than as presented.

@@

\subsubsection{MAC address randomization}

To address the last issue, {\it MAC address randomization}, we rely once again on the fact that we have a lot of data. Namely, we know from the structure of the MAC address whether it has been randomized or not. Figure \ref{fig:nonrandomizedadresses} shows numbers of non-randomized and randomized adresses observed per minute from midnight until around 6:00 am. We can see that their ratio is quite stable through time; in fact, the Pearson correlation coefficient between the time series of randomized and non-randomized addresses is 0.98.

\begin{figure}[h!]
	\centering
	\includegraphics[width=100mm]{rand_nonrand.png}
	\caption{Non-randomized (up) and randomized (down) addresses observed per minute}
	\label{fig:nonrandomizedadresses}
\end{figure}

\begin{figure}[h!]
	\centering
	\includegraphics[width=100mm]{ratio_hist.png}
	\caption{Distribution of ratios between randomized and non-randomized addresses observed per minute}
	\label{fig:ratio_dist}
\end{figure}

\noindent Thus, when estimating crowd density, we ignore the MAC devices that have a randomized address and at the end we multiply the crowd density by a factor, to account for the ignored MAC devices. We choose factor 1.22 to allow for a safety margin: we would rather overestimate than underestimate the crowd density. The ratio  0.22 is the meam plus the standard deviation of the distribution of ratios between randomized and non-randomized addresses observed per minute (Fig \ref{fig:ratio_dist}), but in general it depends on how much underestimation/overestimation is desired. Note also that this ratio should be re-computed periodically, to account for the changing conditions at the smart phones market.

@@

Make an acknowledgments section (Arena, Jessy or maybe Jessy co-author?)

Mention that data is not available under law restrictions

Publish code and cite?
Also data analysis code

explain exactly how the anonymization and data collection works
